---
title: "Case Study 1 DDS"
author: "Avery"
date: "2026-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Executive Summary
```{r}
#Our question of interest we wanted to answer was what were the top factors contributing to employee attrition at Frito Lay. Based on my analysis, I found overtime, job satisfaction, job level, age and monthly income to be the top key factors that were significantly related to attrition. The initial KNN and Naive Bayes models I built to predict attrition had low sensitivity rates, so I balanced the data set from Frito Lay and reconfigured both models to give me a desired sensitivity rate of at least 60%. The final takeaway I got from my analysis was that younger people in lower job level roles seem most likely to leave Frito Lay.

```


#Introduction & Objective
```{r}
#We were given a data set from Frito Lay and asked to identify the key factors driving employee attrition at the company.

```


#Importing the Libraries
```{r}
library(tidyverse)
library(maps)
library(gridExtra)
library(class)
library(caret)
library(e1071)
```


# Import Attrition CSV file
```{r}
#Import CaseStudy1-data.csv
turnover <- read.csv("../assets/data/CaseStudy1-data.csv", header = TRUE)

sum(is.na(turnover))
#No NA values

#Attrition Column Values: "Yes" = The employee left the company, "No" = The employee stayed at the company or is still employed
```


#Plot Summary of Attrition in Percentages
```{r}
#turning Attrition into a factor
turnover$Attrition <- as.factor(turnover$Attrition)
table(turnover$Attrition)
#Attrition is relatively low at 16%

library(ggplot2)
ggplot(data = turnover, aes(x = Attrition, fill = Attrition)) + geom_bar(aes(y = after_stat(count / sum(count)))) + geom_text(stat = "count", aes(y = after_stat(count / sum(count)), label = scales::percent(after_stat(count / sum(count)), accuracy = 0.1)), vjust = -0.3, color = "black") + scale_y_continuous(labels = scales::percent_format()) + labs(title = "Attrition Rate at Frito Lay", x = "Attrition", y = "Percentage of Employees", fill = NULL) + theme(legend.position = "none")

```


#Exploratory Data Analysis and Visualization of Factors
```{r}
colnames(turnover)

#Graph 1 - Monthly Income Histogram
ggplot(data = turnover, aes(x = MonthlyIncome, fill = Attrition)) + geom_histogram() + labs(title = "Attrition by Monthly Income", x = "Monthly Income", y = "Number of Employees") + facet_grid(~Attrition)
#Most employees who left reported a monthly income of lower than $10000

#Graph 2 - Age Density Plot
ggplot(data = turnover, aes(x = Age, fill = Attrition)) + geom_density(alpha = 0.5) + labs(title = "Age Distribution by Attrition", x = "Age", y = "Density")
#More younger employees are leaving the company than older employees

#Graph 3 - Age x Job Level Histograms
ggplot(data = turnover, aes(x = Age, fill = JobLevel)) + geom_histogram() + labs(title = "Distribution of Ages by Job Level", x = "Age", y = "Count", fill = "Job Level") + facet_wrap(~JobLevel) + theme(legend.position = "none")
#I wanted to see if the assertion that more younger people had lower job levels because they had less experience was supported

#Creating a new data set with only 'Yes' for Attrition
turnover_yes <- turnover %>% filter(Attrition == "Yes")
dim(turnover_yes)

#Graph 4 - Job Level (only for those who left)
ggplot(data = turnover_yes, aes(x = JobLevel, fill = Attrition)) + geom_bar(aes(y = after_stat(count / sum(count))), position = "dodge") + geom_text(stat = "count", aes(y = after_stat(count / sum(count)), label = scales::percent(after_stat(count / sum(count)), accuracy = 0.1)), position = position_dodge(width = 0.9), vjust = -0.3, color = "black") + scale_fill_manual(values = c("Yes" = "#00BFC4")) + scale_y_continuous(labels = scales::percent_format()) + labs(title = "Distribution of Employees Who Left By Job Level", x = "Job Level", y = "Percentage of Employees") + theme(legend.position = "none")

#Creating new data set for only job level 1
turnover_level1 <- turnover %>% filter(JobLevel == 1)
dim(turnover_level1)

#Graph 5 - Job Satisfaction x Attrition in Job Level 1
ggplot(data = turnover_level1, aes(x = JobSatisfaction, fill = Attrition)) + geom_bar(aes(y = after_stat(count / sum(count))), position = "dodge") + geom_text(stat = "count", aes(y = after_stat(count / sum(count)), label = scales::percent(after_stat(count / sum(count)), accuracy = 1)), position = position_dodge(width = 0.9), vjust = -0.3, color = "black") + scale_y_continuous(labels = scales::percent_format()) + labs(title = "Attrition by Job Satisfaction in Job Level 1", x = "Job Satisfaction", y = "Percentage of Employees")
#Lower level employees seem to have lower job satisfaction

#Graph 6 - Overtime x Attrition in Job Level 1
ggplot(data = turnover_level1, aes(x = OverTime, fill = Attrition)) + geom_bar(aes(y = after_stat(count / sum(count))), position = "dodge") + geom_text(stat = "count", aes(y = after_stat(count / sum(count)), label = scales::percent(after_stat(count / sum(count)), accuracy = 1)), position = position_dodge(width = 0.9), vjust = -0.3, color = "black") + scale_y_continuous(labels = scales::percent_format()) + labs(title = "Attrition by Overtime in Job Level 1", x = "Overtime", y = "Percentage of Employees") 
#Lower level employees who worked overtime are likely to leave the company more than those who didn't work overtime

```


#KNN Model
```{r}
knn_data <- turnover[, c("Attrition", "JobLevel", "JobSatisfaction", "OverTime", "Age", "MonthlyIncome")]
  
#Predictor variables must be numeric, so we need to convert OverTime to binary variables (1 = Yes, 0 =)
knn_data$OverTime <- ifelse(knn_data$OverTime == "Yes", 1, 0)

#Train / Test split
set.seed(123)

trainIndices <- sample(1:nrow(knn_data), 0.7*nrow(knn_data))

train <- knn_data[trainIndices,]
test <- knn_data[-trainIndices,]

#Scale
train_x <- scale(train[, c("JobLevel", "JobSatisfaction", "OverTime", "Age", "MonthlyIncome")])

train_center <- attr(train_x, "scaled:center") #attr() gets a specific attribute (center) from an object (data frame)
train_scale <- attr(train_x, "scaled:scale")

test_x <- scale(test[, c("JobLevel", "JobSatisfaction", "OverTime", "Age", "MonthlyIncome")], center = train_center, scale = train_scale)

train_y <- train$Attrition
test_y <- test$Attrition

classifications = knn(train_x, test_x, train_y, k = 7)

cm = confusionMatrix(table(Predictions = classifications, Actual = test_y), positive = "Yes")
cm

#Recode Yes/No to Left/Stayed
knn_cm <- as.data.frame(cm$table)
knn_cm <- knn_cm %>% dplyr::mutate(Actual = dplyr::recode(Actual, "Yes" = "Left", "No" = "Stayed"), Predictions = dplyr::recode(Predictions, "Yes" = "Left", "No" = "Stayed"))

#Plotting heatmap
ggplot(knn_cm, aes(x = Predictions, y = Actual, fill = Freq)) + geom_tile() + geom_text(aes(label = Freq)) + scale_fill_gradient(low = "white", high = "blue") + labs(title = "KNN Confusion Matrix", fill = "Frequency") + theme(legend.position = "none")
```


#Determining which k for KNN is best
```{r}
set.seed(123)

trainIndices <- sample(1:nrow(knn_data), 0.7*nrow(knn_data))

train <- knn_data[trainIndices,]
test <- knn_data[-trainIndices,]

#Scale
train_x <- scale(train[, c("JobLevel", "JobSatisfaction", "OverTime", "Age", "MonthlyIncome")])

train_center <- attr(train_x, "scaled:center")
train_scale <- attr(train_x, "scaled:scale")

test_x <- scale(test[, c("JobLevel", "JobSatisfaction", "OverTime", "Age", "MonthlyIncome")], center = train_center, scale = train_scale)

accs = data.frame(accuracy = numeric(50), k = numeric(50))

for(i in 1:50)
{
  classifications = knn(train_x, test_x, train$Attrition, k = i)
  CM = confusionMatrix(classifications, test$Attrition)
  accs$accuracy[i] = CM$overall[1]
  accs$k[i] = i
}

plot(accs$k,accs$accuracy, type = "l", xlab = "k", ylab = "Accuracy", main = "Accuracy vs K")
abline(v = c(7), col = "red")

#Accuracy has the highest steep between k = 1 and k = 10
```


#NB Model
```{r}
nb_data <- turnover[, c("Attrition", "JobLevel", "JobSatisfaction", "OverTime", "Age", "MonthlyIncome")]

#Converting char variables to factors
nb_data$OverTime <- as.factor(nb_data$OverTime)
nb_data$JobLevel <- as.factor(nb_data$JobLevel)

#Train / Test Split
set.seed(123)

trainIndices_nb <- sample(1:nrow(nb_data), 0.7*nrow(nb_data))

train_nb <- nb_data[trainIndices_nb,]
test_nb <- nb_data[-trainIndices_nb,]

#Naive Bayes Model
nb_model <- naiveBayes(Attrition ~ ., data = train_nb, laplace = 1)

nb_class <- predict(nb_model, test_nb)

cm_nb <- confusionMatrix(nb_class, test_nb$Attrition, positive = "Yes")
cm_nb

#Recoding Yes/No as Left/Stayed
nb_cm <- as.data.frame(cm_nb$table)
nb_cm <- nb_cm %>% dplyr::mutate(Reference = dplyr::recode(Reference, "Yes" = "Left", "No" = "Stayed"), Prediction = dplyr::recode(Prediction, "Yes" = "Left", "No" = "Stayed"))

#Plotting heatmap
ggplot(nb_cm, aes(x = Prediction, y = Reference, fill = Freq)) + geom_tile() + geom_text(aes(label = Freq)) + scale_fill_gradient(low = "white", high = "blue") + labs(title = "Naive Bayes Confusion Matrix", fill = "Frequency", x = "Predictions", y = "Actual") + theme(legend.position = "none")
```


#Plotting Sensitivity and Specificity
```{r}
metrics <- data.frame(Model = rep(c("KNN", "Naive Bayes"), each = 2), Metric = rep(c("Sensitivity", "Specificity"), times = 2), Values = c(0.2692, 0.9617,  0.32692, 0.97129))

ggplot(metrics, aes(x = Metric, y = Values, fill = Model)) + geom_bar(stat = "identity", position = "dodge") + geom_text(aes(label = round(Values, 2)), position = position_dodge(width = 0.9), vjust = -0.3) + scale_fill_manual(values = c("KNN" = "#2CA02C", "Naive Bayes" = "#FF7F0E")) + labs(title = "Model Performance Comparison", y = "Score") + theme_minimal()
```


#Reconfiguring KNN model to achieve 60% sensitivity and specificity
```{r}
prop.table(table(knn_data$Attrition))
#The data is imbalanced, with No being at 84% and Yes at 16%

knn_data <- turnover[, c("Attrition", "JobLevel", "JobSatisfaction", "OverTime", "Age", "MonthlyIncome")]
  
#Convert OverTime to numeric
knn_data$OverTime <- ifelse(knn_data$OverTime == "Yes", 1, 0)

#Train / Test split
set.seed(123)

trainIndices <- sample(1:nrow(knn_data), 0.7*nrow(knn_data))

train <- knn_data[trainIndices,]
test <- knn_data[-trainIndices,]

#Partial upsampling to balance out classes
train_yes <- train %>% filter(Attrition == "Yes")
train_no <- train %>% filter(Attrition == "No")

#Upsample (address class imbalance) the "Yes" to 70% of the "No" size
set.seed(123)
train_yes_bal <- train_yes[sample(1:nrow(train_yes), round(0.7*nrow(train_no)), replace = TRUE), ]

#Combine balanced dataset
train_balanced <- bind_rows(train_no, train_yes_bal)

#Separate predictors and outcomes
train_x <- scale(train_balanced[, c("JobLevel", "JobSatisfaction", "OverTime", "Age", "MonthlyIncome")])
train_y <- train_balanced$Attrition

#Scaling Parameters
train_center <- attr(train_x, "scaled:center") #attr() gets a specific attribute (center) from an object (data frame)
train_scale <- attr(train_x, "scaled:scale")

test_x <- scale(test[, c("JobLevel", "JobSatisfaction", "OverTime", "Age", "MonthlyIncome")], center = train_center, scale = train_scale)
test_y <- test$Attrition

#Determining the strongest k-value
k_values <- c(3, 5, 7, 9)
knn_results <- data.frame(k = integer(), Sensitivity = numeric(), Specificity = numeric(), Accuracy = numeric())

for (k in k_values) {
  predictor <- knn(train_x, test_x, train_y, k = k)
  cm <- confusionMatrix(table(Predictions = predictor, Actual = test_y), positive = "Yes")
  
  knn_results <- rbind(knn_results, data.frame(
    k = k,
    Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
    Specificity = as.numeric(cm$byClass["Specificity"]),
    Accuracy = as.numeric(cm$overall["Accuracy"])))
}

best_k <- knn_results %>%
  filter(Sensitivity >= 0.6, Specificity >= 0.6) %>%
  arrange(desc(Accuracy)) %>%
  slice(1)

best_k
#Best k = 3

#Final model
final_classifications = knn(train_x, test_x, train_y, k = best_k$k)

final_cm = confusionMatrix(table(Predictions = final_classifications, Actual = test_y), positive = "Yes")
final_cm

#Plotting KNN heatmap
cm_final <- as.data.frame(final_cm$table)
cm_final <- cm_final %>% dplyr::mutate(Actual = dplyr::recode(Actual, "Yes" = "Left", "No" = "Stayed"), Predictions = dplyr::recode(Predictions, "Yes" = "Left", "No" = "Stayed"))

ggplot(cm_final, aes(Predictions, Actual, fill = Freq)) + geom_tile() + geom_text(aes(label = Freq)) + scale_fill_gradient(low = "white", high = "blue") + labs(title = "KNN Confusion Matrix", fill = "Frequency") + theme(legend.position = "none")
```


#Reconfiguring NB Model to achieve at least 60% sensitivity and specificity
```{r}
nb_data <- turnover[, c("Attrition", "JobLevel", "JobSatisfaction", "OverTime", "Age", "MonthlyIncome")]

#Converting char variables to factors
nb_data$OverTime <- as.factor(nb_data$OverTime)
nb_data$JobLevel <- as.factor(nb_data$JobLevel)

#Train / Test Split
set.seed(123)

trainIndices_nb <- sample(1:nrow(nb_data), 0.7*nrow(nb_data))

train_nb <- nb_data[trainIndices_nb,]
test_nb <- nb_data[-trainIndices_nb,]

#Balancing the Data
train_nb_yes <- train_nb %>% filter(Attrition == "Yes")
train_nb_no <- train_nb %>% filter(Attrition == "No")

#Upsample (address class imbalance) the "Yes" to 70% of the "No" size
set.seed(123)
train_yes_nb_bal <- train_nb_yes[sample(1:nrow(train_nb_yes), round(0.7*nrow(train_nb_no)), replace = TRUE), ]

#Combine balanced dataset
train_nb_balanced <- bind_rows(train_nb_no, train_yes_nb_bal)

#Naive Bayes Model
nb_model <- naiveBayes(Attrition ~ ., data = train_nb_balanced, laplace = 1) 

#Predicting probabilities
nb_class <- predict(nb_model, test_nb, type = "raw")[, "Yes"]

#Applying a threshold
threshold <- 0.45
nb_class_new <- factor(ifelse(nb_class > threshold, "Yes", "No"), levels = c("No", "Yes"))

#Confusion matrix with threshold
cm_nb_new <- confusionMatrix(nb_class_new, test_nb$Attrition, positive = "Yes")
cm_nb_new

#Plotting new heatmap
nb_cm_new <- as.data.frame(cm_nb_new$table)
nb_cm_new <- nb_cm_new %>% dplyr::mutate(Reference = dplyr::recode(Reference, "Yes" = "Left", "No" = "Stayed"), Prediction = dplyr::recode(Prediction, "Yes" = "Left", "No" = "Stayed"))

ggplot(nb_cm_new, aes(Prediction, Reference, fill = Freq)) + geom_tile() + geom_text(aes(label = Freq)) + scale_fill_gradient(low = "white", high = "blue") + labs(title = "Naive Bayes Confusion Matrix", fill = "Frequency", x = "Predictions", y = "Actual") + theme(legend.position = "none")
```


#Plotting Sensitivity and Specificity for New Models
```{r}
metrics_new <- data.frame(Model = rep(c("KNN", "Naive Bayes"), each = 2), Metric = rep(c("Sensitivity", "Specificity"), times = 2), Values = c(0.6154, 0.7225,  0.6538, 0.6459))

ggplot(metrics_new, aes(x = Metric, y = Values, fill = Model)) + geom_bar(stat = "identity", position = "dodge") + geom_text(aes(label = round(Values, 2)), position = position_dodge(width = 0.9), vjust = -0.3) + scale_fill_manual(values = c("KNN" = "#2CA02C", "Naive Bayes" = "#FF7F0E")) + labs(title = "New Model Performance Comparison", y = "Score") + theme_minimal()
```


#Validation Portion - No Attrition CSV
```{r}
#Import CaseStudy1CompSet No Attrition.csv
competition <- read.csv("../assets/data/CaseStudy1CompSet No Attrition.csv", header = TRUE)
colnames(competition)

#Convert overtime into a factor to be a predictor
competition$OverTime <- ifelse(competition$OverTime == "Yes", 1, 0)
competition_data <- competition[, c("JobLevel", "JobSatisfaction", "OverTime", "Age", "MonthlyIncome")]

#Scale using training parameters 
competition_data <- scale(
  competition_data,
  center = train_center,
  scale = train_scale
)

competition_predictions <- knn(
  train = train_x,
  test = competition_data,
  cl = train_y,
  k = best_k$k
)

length(competition_predictions) #300

#Creating a CSV file
competition_submission <- data.frame(
  ID = competition$ID,
  Attrition = competition_predictions
)

#Convert predictions into characters to prevent formatting issues
competition_submission$Attrition <- as.character(competition_submission$Attrition)

#Order by ID
competition_submission <- competition_submission[order(competition_submission$ID), ]

#Exporting the CSV file
write.csv(
  competition_submission,
  "Case1PredictionsJohnAttrition.csv",
  row.names = FALSE
)
```


#Predictions
```{r}
predictions <- read.csv("../assets/data/Case1PredictionsJohnAttrition.csv")
head(predictions)
```


# Video Presentation

Click here to watch my presentation: [My YouTube Video](https://youtu.be/IEmBZYtSBw0)